---
format:
  gfm:
    default-image-extension: ""
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Introduction

`Spatialdata.data` package provides utilities for accessing, reading and 
generating SpatialData datasets. 

*scverse* SpatialData examples are available through

1. Bioc's NSF OSN bucket and 
2. scverse's spatialdata-sandbox ([https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html](https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html))

which are accessible from within R, using `BiocFileCache`.

`SpatialData.data` uses `basilisk` to interface and maintain multiple 
versions of python's spatialdata package (0.3.0, 0.5.0 and 0.7.2) for reading
and writing to .zarr packages. 

`basilisk` environments (only 0.5.0 and 0.7.2) are also accompanied by the
`dummy-spatialdata` package that generates toy spatialdata examples 
whose elements are customized by the user.

# Installation

```{r install, eval=FALSE}
#| label: install
if(!requireNamespace("BiocManager"))
  install.packages("BiocManager")
BiocManager::install("HelenaLC/spatialdataR")
BiocManager::install("HelenaLC/SpatialData.data")
```

```{r library}
library(spatialdataR)
library(SpatialData.data)
```

To *interrogate* our S3 bucket you will need [paws](https://cran.r-project.org/web/packages/paws/index.html) 
installed; it is not necessary for retrievals.

```{r aws}
if(!requireNamespace("paws"))
  install.packages("paws")
library(paws)
Sys.setenv(AWS_REGION = "us-east-1") 
```

## Load SpatialData (.zarr) from Archives

Data from a variety of technologies has been made available as `SpatialData` .zarr stores 
[here](https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html).
These, in turn, have been deposited in Bioconductor's NSF Open Storage Network also and 
can be retrieved with caching support using `r BiocStyle::Biocpkg("BiocFileCache")`.

Any spatialdata dataset can be retrieved (once) into some location, and 
read into R.

```{r load}
#| label: load
(x <- SD.data_load("ColorectalCarcinomaMIBITOF"))
```

## SpatialData-io

`SpatialData.data` also provides access to some raw spatial omic readouts. These
data bundle can then be converted into SpatialData objects using `spatialdata-io`
pyton package via `basilisk`.

```{r avail}
#| label: avail
SD.data_available("biocOSN_Xenium")
```

We use `options(sd_version)` to set the SpatialData version.

```{r load_basilisk}
#| label: load_basilisk
options(sd_version = "0.3.0")
(x <- SD.data_load("Breast2fov_10x", source = "biocOSN_Xenium"))
```

## Dummy SpatialData objects

`SpatialData.data` package incorporates the `dummy-spatialdata` python package
from PyPI ([https://pypi.org/project/dummy-spatialdata/](https://pypi.org/project/dummy-spatialdata/))
via `basilisk` to generate toy spatialdata objects in multiple spatialdata 
versions. 

```{r generate_dataset}
#| label: generate_dataset
zarrfile <- tempfile(fileext = ".zarr")
generate_dataset(
  file = zarrfile, 
  sd_version = "0.5.0",
  images = list(
    list(type = "rgb", scale_factors = c(2L,2L,2L), coordinate_system="global"),
    list(type = "grayscale", n_layers = c(), coordinate_system="global")
  ),
  shapes = list(
    list(n=12L, type="polygon", coordinate_system="global")
  ),
  points = list(
    list(n=12L)
  ),
  coordinate_systems = list(
    global = list(
      transformations = list("affine"), 
      shape = list(x=2000L, y=2000L)
    )
  )
)
sd <- spatialdataR::readSpatialData(zarrfile)
```