---
format:
  gfm:
    default-image-extension: ""
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Introduction

`SpatialData.data` package provides utilities for accessing, reading and generating SpatialData datasets. Data from a variety 
of spatial omics technologies has been made available as `SpatialData` (zipped) .zarr stores.

These *scverse* SpatialData examples are available through sources 

1.  **biocOSN:** Bioc’s NSF OSN bucket,
2.  **biocOSN_Xenium:** Bioc’s NSF OSN bucket for raw data outputs from some Xenium datasets and
3.  **sandbox:** scverse’s spatialdata-sandbox
    (<https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html>)

`SpatialData.data` uses `basilisk` to interface and maintain multiple versions of scverse’s `spatialdata` module (0.5 and 0.8) for reading and writing to .zarr stores. 

The package also incorporates `dummy-spatialdata` python module that generates toy SpatialData examples whose elements are customized by the user. These examples can be generated again using `spatialdata` module versions 0.5 and 0.8. 

Please visit the **vignette** for more information.

# Installation

```{r install, eval=FALSE}
#| label: install
if(!requireNamespace("BiocManager"))
  install.packages("BiocManager")
BiocManager::install("spatialdataR")
BiocManager::install("SpatialData.data")
```

```{r library}
library(spatialdataR)
library(SpatialData.data)
```

To *interrogate* our S3 bucket you will need [paws.storage](https://cran.r-project.org/web/packages/paws.storage/index.html) 
installed. 

```{r aws}
if(!requireNamespace("paws.storage"))
  install.packages("paws.storage")
library(paws.storage)
Sys.setenv(AWS_REGION = "us-east-1") 
```

## Load SpatialData (.zarr) from Archives

Any spatialdata dataset can be retrieved (once) into some location, and
read into R.

```{r load}
#| label: load
(x <- SD.data_load("ColorectalCarcinomaMIBITOF"))
```

You can view a list of available datasets using:

```{r load-data-list}
SD.data_list()
```