---
title: "`SpatialData.data`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData.data')`"
author: 
  - name: Vince Carey
  - name: Helena Lucia Crowell
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{SpatialData.data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

`r BiocStyle::Biocpkg("SpatialData.data")` package provides utilities for accessing, reading and generating SpatialData datasets. Data from a variety 
of spatial omics technologies has been made available as `SpatialData` (zipped) .zarr stores.

These *scverse* SpatialData examples are available through sources 

1.  **biocOSN:** Bioc’s NSF OSN bucket,
2.  **biocOSN_Xenium:** Bioc’s NSF OSN bucket for raw data outputs from some Xenium datasets and
3.  **sandbox:** scverse’s spatialdata-sandbox
    (<https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html>)

`r BiocStyle::Biocpkg("SpatialData.data")` uses `basilisk` to interface and maintain multiple versions of scverse’s `spatialdata` module (0.5 and 0.8) for
reading and writing to .zarr stores.

The package also incorporates `dummy-spatialdata` python module that generates toy SpatialData examples whose elements are customized by the user. These examples can be generated again using `spatialdata` module versions 0.5 and 0.8

## Installation

You can install `r BiocStyle::Biocpkg("SpatialData.data")` using: 

```{r eval=FALSE}
if(!requireNamespace("spatialdataR"))
  BiocManager::install("spatialdataR")
if(!requireNamespace("SpatialData.data"))
  BiocManager::install("SpatialData.data")
```

You can also install the development version like so:

```{r install-pak, eval=FALSE}
if(!requireNamespace("pak"))
  install.packages("pak")
pak::pak("HelenaLC/SpatialData.data")
```

To *interrogate* our S3 bucket you will need [paws.storage](https://cran.r-project.org/web/packages/paws.storage/index.html) 
installed.

```{r load-libs}
library(spatialdataR)
library(SpatialData.data)
library(paws.storage)
Sys.setenv(AWS_REGION = "us-east-1") 
```

## Load SpatialData (.zarr) from Archives

Any SpatialData dataset can be retrieved (once) into some location, and
read into R.

```{r load-data}
(x <- SD.data_load("ColorectalCarcinomaMIBITOF"))
```

You can also install the same data from different sources, including the scverse's `spatialdata` sandbox where SpatialData stores are saved as Zarr v3.

```{r load-data-sandbox}
(x <- SD.data_load("ColorectalCarcinomaMIBITOF", source = "sandbox"))
```

We can check all available datasets and their sources with:

```{r load-data-list}
SD.data_list()
```

or as below for a detailed overview and metadata on all datasets: 

```{r view-data-list, eval=FALSE}
View(SD.data_list(metadata = TRUE))
```

You can also interrogate the sources (S3 buckets) for available (zipped) .zarr archives:

```{r avail}
SD.data_available("biocOSN")
```

## Using spatialdata-io

`SpatialData.data` also provides access to some raw spatial omic
readouts. After installation from the source (i.e. `biocOSN_Xenium`), 
these data bundles can then be converted into SpatialData
objects using `spatialdata-io` python package.

You can use `basilisk` to convert these readouts into SpatialData zarr stores 
using multiple `spatialdata` module versions, each associated with a different 
Zarr format:

* **0.5.0** (Zarr v2) and 
* **0.8.0** (Zarr v3)

We use `options(sd_version)` to set the `spatialdata` module version.

```{r sd-io}
options(sd_version = "0.5.0")
(x <- SD.data_load("Breast2fov_10x", source = "biocOSN_Xenium"))
```

## Generating dummy SpatialData objects

`SpatialData.data` package incorporates the `dummy-spatialdata` python package ([https://pypi.org/project/dummy-spatialdata/](https://pypi.org/project/dummy-spatialdata/))
via `basilisk` to generate toy spatialdata objects in multiple spatialdata 
versions. 

```{r dummy}
sd_zarr <- generate_dataset(
  sd_version = "0.5.0",
  images = list(
    list(type = "rgb", scale_factors = c(2L,2L,2L), coordinate_system="global"),
    list(type = "grayscale", n_layers = c(), coordinate_system="global")
  ),
  shapes = list(
    list(n=12L, type="polygon", coordinate_system="global")
  ),
  points = list(
    list(n=12L)
  ),
  coordinate_systems = list(
    global = list(
      transformations = list("affine"), 
      shape = list(x=2000L, y=2000L)
    )
  )
)
sd_zarr
```

Now we can read the SpatialData object with SpatialData.

```{r read}
sd <- readSpatialData(sd_zarr)
sd
```

We can also get individual elements

```{r image}
image(sd, 1)
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
