---
title: "`SpatialData.data`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData.data')`"
author: 
  - name: Vince Carey
  - name: Helena Lucia Crowell
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{SpatialData.data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

```{r load-libs}
library(SpatialData)
library(SpatialData.data)
library(paws)
```

`Spatialdata.data` package provides utilities for accessing, reading and 
generating SpatialData datasets. 

*scverse* SpatialData examples are available through

1. Bioc's NSF OSN bucket and 
2. scverse's spatialdata-sandbox ([https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html](https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html))

which are accessible from within R, using `BiocFileCache`.

`SpatialData.data` uses `basilisk` to interface and maintain multiple 
versions of python's spatialdata package (0.3.0, 0.5.0 and 0.7.2) for reading
and writing to .zarr packages. 

`basilisk` environments (only 0.5.0 and 0.7.2) are also accompanied by the
`dummy-spatialdata` package that generates toy spatialdata examples 
whose elements are customized by the user.

# Introduction

```{r aws}
Sys.setenv(AWS_REGION = "us-east-1")
```

## Zipped .zarr archives

Data from a variety of technologies has been made available as `SpatialData` .zarr stores 
[here](https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html).
These, in turn, have been deposited in Bioconductor's NSF Open Storage Network also and 
can be retrieved with caching support using `r BiocStyle::Biocpkg("BiocFileCache")`.

Any spatialdata dataset can be retrieved (once) into some location, and 
read into R.  We use dataset-specific functions, or `load_data`:

```{r load-data}
(x <- load_data("ColorectalCarcinomaMIBITOF")) # stub can be used
```

or 

```{r load-data-biocOSN}
# from biocOSN
x <- ColorectalCarcinomaMIBITOF()
```

```{r load-data-sandbox}
# TODO: zarr v3 is not complete yet
# from sandbox (Zarr v3)
# x <- ColorectalCarcinomaMIBITOF(source = bucket_path("sandbox"))
```

We can check all available datasets below:

```{r load-data-list}
SpatialData.data_list()
```

To interrogate the bucket for available (zipped) .zarr archives:

```{r avail}
available("biocOSN")
```

## SpatialData-io

`SpatialData.data` also provides access to some raw spatial omic readouts. These
data bundle can then be converted into SpatialData objects using `spatialdata-io`
pyton package via `basilisk`.

```{r avail2, eval=FALSE}
available("biocOSN_Xenium")
```

You can use `basilisk` to convert these readouts into various SpatialData 
formats: 

* **0.3.0** (Zarr v2), 
* **0.5.0** (Zarr v2) and 
* **0.7.2** (Zarr v3)

We use `options(sd_version)` to set the SpatialData version.

```{r sd-io, eval=FALSE}
options(sd_version = "0.3.0")
(x <- Breast2fov_10x(source = bucket_path("biocOSN_Xenium")))
```

## Dummy SpatialData objects

`SpatialData.data` package incorporates the `dummy-spatialdata` python package
from PyPI ([https://pypi.org/project/dummy-spatialdata/](https://pypi.org/project/dummy-spatialdata/))
via `basilisk` to generate toy spatialdata objects in multiple spatialdata 
versions. 

```{r dummy, eval=FALSE}
sd <- generate_dataset(
  sd_version = "0.5.0",
  images = list(
    list(type = "rgb", n_layers = 4L, coordinate_system="global"),
    list(type = "grayscale", n_layers = 1L, coordinate_system="global")
  ),
  shapes = list(
    list(n_shapes=12L, coordinate_system="global")
  ),
  points = list(
    list(n_points=12L)
  ),
  coordinate_systems = list(
    global = list(
      transformations = list("affine"), 
      shape = list(x=2000L, y=2000L)
    )
  )
)
readSpatialData(sd)
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
