---
title: "`SpatialData.data`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData.data')`"
author: 
  - name: Vince Carey
  - name: Helena Lucia Crowell
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{SpatialData.data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

```{r load-libs}
library(SpatialData)
library(SpatialData.data)
```

# Introduction

Data from a variety of technologies has been made available as `SpatialData` .zarr stores 
[here](https://spatialdata.scverse.org/en/latest/tutorials/notebooks/datasets/README.html).
These, in turn, have been deposited in Bioconductor's NSF Open Storage Network bucket,
and can be retrieved with caching support using `r BiocStyle::Biocpkg("BiocFileCache")`.

We can interrogate the bucket for available (zipped) .zarr archives:

```{r ons-list}
if (requireNamespace("paws")) availableOSN()
```

Any of the above can be retrieved (once) into some location, and read into R.  We
use dataset-specific functions, or `loadFromOSN`:

```{r ons-read}
ls("package:SpatialData.data")  # run a dataset-specific function, or
(x <- loadFromOSN("Lung2fov")) # stub can be used
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
